Formal Grammars as Types in Non-commutative Linear-Non-Linear Type Theory

### 2024-07-08 POPL/paper.tex
intro to syntax, intro editing, citations

\setcounter{page}{1}

\section{Introduction}

\section{A Syntax-Independent Notion of Syntax}
The notion of formal language is central to the theory of parsing. A
\emph{formal language} $L$ over an alphabet $\Sigma$ is classically defined as a
\emph{subset} of strings $L \subseteq \String$. This definition is
especially useful as it gives a semantics to formal grammars that is completely
independent of any particular syntactic grammar formalism. Any new
notion of grammar can be given a language semantics and it provides a precise
mathematical specification for implementing a \emph{recognizer} of a language.

That is, a recognizer should be extensionally equivalent to the indicator function
$\chi_{L} : \String \to \Prop$ mapping each string $w$ to the the proposition
that it is in the language $w \in L$. However, language recognition is
insufficient for specifying a parser --- a function that decomposes a string
into semantic components that adhere to the structure of a formal grammar. In
practice, parsers do more than simply return whether a string is valid or
invalid; instead, they emit structured \emph{parse trees}\footnote{Such parse
  trees are usually not materialized in memory, as semantic actions often perform
  computation over the trees while they are being constructed.}

@@

As they a simply sets, formal languages are naturally endowed with a partial
order via subset inclusion. Similarly, formal grammars are naturally endowed
with the structure of a \emph{category}. A morphism $f : G \to H$ between two
grammars is given by a family of functions

@@

The category-theoretic framework can be further used to describe and compare
different notions of formal grammar. For instance, Chomskyan generative
grammars, Thue systems, Montague grammars, and so on are all distinct syntactic
presentations of the same underlying idea of an abstract specification for
parsing. Each of these is an instance of a general \emph{notion of formal
  grammar} --- a category paired with a functor into
$\Set^{\String}$. That is, they may all be interpreted in our category of grammars. In this sense, $\Set^{\String}$ provides a universal semantic
domain to reason about parsing. The slice 2-category over $\Set^{\String}$ is
then the appropriate setting to reason over notions of formal grammar.

@@

Our contributions are then:

\steven{These sound awkward but I'm painting with broad strokes rn}
\begin{itemize}
  \item A syntax-independent definition of formal grammar
  \item A dependent
    linear/non-linear (LNL) type theory to syntactically capture the semantics of parsing
  \item A categorification of Kleene Algebra
  \item Interpretations of the LNL type system for reasoning about formal grammars and formal languages
  \item The presentation of automata as grammars. By internalizing automata ---
        the operational view of a parser --- we may extract executable parsers
  \item A language implemented shallowly in Agda
\end{itemize}

\section{Categorification of Formal Grammars}
\steven{Idk if this section has the right title, but I wanted to change it if we are adding ``chomsky categories''}

@@

This result feels similar in spirit to the definition of action
algebras, which are algebras where the product also has adjoint
operations which results in the Kleene star being more easily
axiomatized \steven{TODO cite}
% \cite{kozen1994}.

@@

\begin{lemma}[TODO cite
  % \cite{day1970}
  ]
  Under the assumptions above, the presheaf category $\Set^{\cat{C}}$ is
  monoidal closed.
\end{lemma}

@@

With this in mind, we want to design a dependent type theory that
builds on top of Kleene categories with the goal of internalizing
various reasoning principles for languages and their parsers. This
leads us to the abstract notion of model we are interested in
capturing with our type theory.

\begin{definition}
  A \emph{Chomsky category} is a locally Cartesian category with two hierarchies of
  universes $\{L_i\}_{i\in \nat}$ and $\{U_i\}_{i\in \nat}$ such that
  every $L_i$ and $U_i$ are $U_{i+1}$-small. Furthermore, we require
  $U_i$ to be closed under dependent products and sum,
  $L_i$ to be closed under the Kleene category connectives,
  dependent products, left and right closed structures, with
  a type constructor $G : L_i \to U_i$ and a linear dependent sum
  going the other direction.
\end{definition}

\begin{theorem}
  The presheaf category $\Set^{\cat{\Sigma^*}}$ is a model.
\end{theorem}

@@

In 1995, Benton et al.
\cite{bentonMixedLinearNonlinear1995} first gave categorical models of a logic
that synthesized linear and non-linear logic. Substructural logics, like linear and separation logic, offer precise control
over how values from a type may be used, in particular how many times a value
may be used in a derivation. This control enables \emph{resource-sensitive}
reasoning, where we may treat a resource as \emph{consumed} after usage. In the
context of parsing, we may treat the characters composing a string as finite resources
that are consumed at parse-time.

Building off of Benton et al., Krishnaswami et al.
\cite{krishnaswami_integrating_2015} developed a type
system that unifies both linear and
dependent types --- referred to as a \emph{linear-non-linear} (LNL) type theory. In
such settings, there are two sub-theories --- one linear and one non-linear. The
linear sub-theory embeds ordinary linear types without dependence, while the
non-linear sub-theory embeds standard Martin-L\"of type theory. The interplay of
linear types with dependence lives at the boundary between these two fragments.
Benton et al.\ gives an accounting of how we may jump between these
fragments via a monoidal adjunction $F \dashv G$ between linear terms and
non-linear terms.

@@

The structrual judgements are shown in \Cref{fig:structjdg}, the typing well-formedness in \Cref{fig:typewf}, and the intuitionistic typing rules in \Cref{fig:inttyping}. There are all just as they appear in \cite{krishnaswami_integrating_2015}.

@@

\begin{figure}
  \begin{mathpar}
    \inferrule{~}{\Gamma \vdash U_i : U_{i+1}}
 %
    \and
%
    \inferrule{~}{\Gamma \vdash L_i : U_{i+1}}
%
    \\
%
    \inferrule{\Gamma \vdash X : U_i \\ \hspace{-0.1cm} \Gamma, x : X \vdash Y : U_i}{\Gamma \vdash \PiTy x X Y : U_i }%
%
    \and
%
    \inferrule{\Gamma\vdash X : U_i \\ \hspace{-0.1cm} \Gamma, x : X \vdash Y : U_i}{\Gamma \vdash \SigTy x X Y : U_i}
%
    \\
%
    \inferrule{~}{\Gamma \vdash 1 : U_i}
%
    \and
%
    \inferrule{\Gamma \vdash A : L_i}{\Gamma \vdash G A : U_i}
%
    \\
%
    \inferrule{~}{\Gamma \vdash I : L_i}
 %
    \and
%
    \inferrule{\Gamma \vdash A : L_i \\ \hspace{-0.1cm}\Gamma \vdash B : L_i}{\Gamma \vdash A \otimes B : L_i}
%
    \and
%
    \inferrule{\Gamma \vdash A : L_i \\ \hspace{-0.1cm}\Gamma \vdash B : L_i}{\Gamma \vdash A \lto B : L_i}
%
    \and
%
    \inferrule{\Gamma \vdash A : L_i \\ \hspace{-0.1cm}\Gamma \vdash B : L_i}{\Gamma \vdash B \tol A : L_i}
%
    \\
%
    \inferrule{\Gamma \vdash X : U_i \\ \Gamma, x : X \vdash A : L_i}{\Gamma \vdash \LinPiTy x X A : L_i}
%
    \and
%
    \inferrule{\Gamma \vdash X : U_i \\ \Gamma, x : X \vdash A : L_i}{\Gamma \vdash \LinSigTy x X A : L_i}
%
    \\
%
    \inferrule{\Gamma \vdash X : U_i \quad \{\Gamma \vdash e_i : X\}_i}{\Gamma \vdash e_1 =_X e_2 : U_i}
    %
    \and
    %
    \inferrule{~}{\Gamma \vdash \top : L_i}
%
    \and
%
    \inferrule{\Gamma \vdash A : L_i \quad \Gamma \vdash B : L_i}{\Gamma \vdash A \amp B : L_i}
%
    \\
    %
    \inferrule{c \in \Sigma}{\Gamma \vdash c : L_0}
    %
    \and
    %
    \inferrule{\Gamma, x : L_i \vdash A : L_i \and A \textrm{ strictly positive}}{\Gamma \vdash \mu x.\, A : L_i}
  \end{mathpar}
  \caption{Type well-formedness}
  \label{fig:typewf}
\end{figure}

@@

    \inferrule{\Gamma \vdash p : e =_X e'}{\Gamma \vdash e \equiv e' : X}
%
    \and
%
    \inferrule{~}{\Gamma \vdash \app {(\lamb x e)} {e'} \equiv \subst x e {e'} : X}
%
    \and
%
    \inferrule{~}{\Gamma \vdash e \equiv \lamb x {\app e x} : \PiTy x X Y}
%
    \and
%
    \inferrule{~}{\Gamma \vdash \pi_1\, (e_1, e_2) \equiv e_1 : X}
%
    \and
%
    \inferrule{~}{\Gamma \vdash \pi_2\, (e_1, e_2) \equiv e_2 : \subst x {e_1} Y}
%
    \and
%
    \inferrule{~}{\Gamma \vdash e \equiv (\pi_1\, e, \pi_2\, e) : \SigTy x X Y}
%
    \and
%
    \inferrule{~}{}
%
    \inferrule{~}{\Gamma \vdash t \equiv t' : 1}
%
    \and
%
    \inferrule{~}{\Gamma \vdash G\, (G^{-1} \, t) \equiv t : G A}
%
    \and

@@

    \\
%
    \inferrule{~}{\Gamma; \cdot \vdash G^{-1}\, (G \, t ) \equiv t: A}
%
    \and
%
    \inferrule{~}{\Gamma; \Delta \vdash \app {(\lamblto a e)} {e'} \equiv \subst e x {e'} : C}
%
    \and
%
    \inferrule{~}{\Gamma; \Delta \vdash e \equiv \lamblto a {\app e a} : A \lto B}
%
    \and
%
    \inferrule{~}{\Gamma; \Delta \vdash \app {(\lambtol a e)} {e'} \equiv \subst e x {e'} : C}
%
    \and
%
    \inferrule{~}{\Gamma; \Delta \vdash e \equiv \lambtol a {\app e a} : A \tol B}
%
    \and
%
    \inferrule{~}{\Gamma; \Delta \vdash \app {(\dlamb x e)} {e'} \equiv \subst x a {e'} : C}
%
    \and
%
    \inferrule{~}{\Gamma; \Delta \vdash e \equiv \dlamb x {\app e x} : \LinPiTy x X A}
%
    \and
%
    \inferrule{~}{\Gamma; \Delta \vdash e \equiv e' : \top}
%
    \and
%
    \inferrule{~}{\Gamma; \Delta \vdash e_i \equiv \pi_i (e_1, e_2) : A_i}
%
    \and
%
    \inferrule{~}{\Gamma; \Delta \vdash e \equiv (\pi_1 e, \pi_2 e) : A\& B}
%
    \and
%
    \inferrule{~}{\Gamma; \Delta \vdash \letin {()} {()} e \equiv e : C}
    %
    \and
    %
    \inferrule{~}{\Gamma; \Delta \vdash \letin {()} e {\subst {e'} {()} a} \equiv \subst {e'} a e : C}
%
    \and
%
    \inferrule{~}{\Gamma; \Delta \vdash \letin {e \otimes e'} {a \otimes a'} e'' \equiv \subst {e''} {a, a'} {e, e'} : C}
%
    \and
    %
    \inferrule{~}{\Gamma; \Delta \vdash \letin {a \otimes b} e {\subst {e'} {a \otimes b} c} \equiv \subst {e'} c e : C}
%
    \and
%
    \inferrule{~}{\Gamma;\Delta \vdash \letin {(x, a)} {(e, e')} {e''} \equiv \subst {e''} {x, a} {e, e'} : C}
    %
    \and
    %
    \inferrule{~}{\Gamma; \Delta \vdash \letin {(x, a)} e {\subst {e'} {(x, a)} y} \equiv \subst {e'} e y : C}
\end{mathpar}
  \caption{Judgmental equality}
  \label{fig:jdgeq}
\end{figure}
\subsection{Syntax}
\label{subsec:syntax}

@@

\subsection{Semantics in Grammars}
\label{subsec:seming}
As mentioned earlier, the interpretation of grammars as functions from
$\String$ to $\Set$ serves as our intended semantics. When designing the syntax of this formalism, we were often
informed by this model. That is, there is an interpretation
$\llbracket \cdot \rrbracket$ of our syntax that provides grammar-based
semantics.

@@

\section{Automata as Grammars}
Classically, formal language theory is closely related to the study of automata.
In the Chomsky hierarchy, each language class is associated to a class of
automata that serve as recognizers. Internal to the type theory of \cref{sec:tt},
we can characterize these language classes syntactically; moreover, we
demonstrate the equivalence of these language classes to its associated automata class as a
proof term within our logic.

@@

\section{Related work}

>>>>

\section{A Syntax-Independent Notion of Syntax}
\label{sec:intro}
The theory of formal languages and parsing is one of the oldest and
most thoroughly developed areas of theoretical computer science. A
prominent topic in the 1950, 60s, and 70s led to a series of remarkable
developments: Chomsky's hierarchy of grammar formalisms \cite{chomThreeModels1956}, practical
algorithms for parsing of regular and context-free grammars and
variants \cite{KNUTH1965607,Earley1970}, as well as implementations of practical tools for the
generation of efficient parsers \cite{Johnsonyacc}.

The notion of formal language is central to the theory of parsing. A
\emph{formal language} $L$ over an alphabet $\Sigma$ is classically defined as a
\emph{subset} of strings $L \subseteq \String$. This definition is
especially useful as it gives a semantics to formal grammars that is completely
independent of any particular syntactic grammar formalism. Any new
notion of grammar can be given a language semantics and it provides a precise
mathematical specification for implementing a \emph{recognizer} of a language.

That is, a recognizer should be extensionally equivalent to the indicator function
$\chi_{L} : \String \to \Prop$ mapping each string $w$ to the the proposition
that it is in the language $w \in L$. However, language recognition is
insufficient for specifying a parser --- a function that decomposes a string
into semantic components that adhere to the structure of a formal grammar. In
practice, parsers do more than simply return whether a string is valid or
invalid; instead, they emit structured \emph{parse trees}\footnote{Such parse
  trees are usually not materialized in memory, as semantic actions often perform
  computation over the trees while they are being constructed.}.

@@

As they are simply sets, formal languages are naturally endowed with a partial
order via subset inclusion. Similarly, formal grammars are naturally endowed
with the structure of a \emph{category}. A morphism $f : G \to H$ between two
grammars is given by a family of functions

@@

The category-theoretic framework can be further used to describe and compare
different notions of formal grammar. For instance, Chomskyan generative
grammars, semi-Thue systems, Montague grammars, and so on are all distinct syntactic
presentations of the same underlying idea of an abstract specification for
parsing. Each of these is an instance of a general \emph{notion of formal
  grammar} --- a category paired with a functor into
$\Set^{\String}$. That is, they may all be interpreted in our category of grammars. In this sense, $\Set^{\String}$ provides a universal setting to reason about parsing.
The slice 2-category over $\Set^{\String}$ then provides the appropriate expressivity to prove equivalences between grammar formalism, such as the equivalence between semi-Thue systems and Chomsky's unresticted grammars.

@@

This paper begins in \cref{sec:categorify} by giving a categorical interpretation of Kleene algebra. However, Kleene algebras are primarily concerned with the regular languages. With this in mind, we further abstract our categorical model from the connectives of regular grammars to those of context-free grammars and beyond. From there, in \cref{sec:tt} we provide a type theory to construct and reason about such formal grammars. In \cref{sec:semantics} we give interpretations of the syntax as grammars and as languages, and prove a canonicity theorem for the type theory. Finally, in \cref{sec:automata} we give a presentation of automata as grammars. With this view in mind, we may encode parsing algorithms and execute them entirely internal to our formalism. Our contributions are then:

\begin{itemize}
  \item A universal semantic notion of formal grammar given by the category $\Set^{\String}$
  \item Categorical structures that capture the concepts integral to formal grammar theory, and that are notably modeled by $\Set^{\String}$
  \item A dependent
    linear/non-linear (LNL) type theory to syntactically capture the above categorical structures
  \item An implementation of the language, which is emedded shallowly in Agda
\end{itemize}

\section{Categorification of Formal Grammars}
\label{sec:categorify}
\steven{Idk if this section has the right title, but I wanted to change it if we are adding ``chomsky categories''}

@@

This result feels similar in spirit to the definition of action
algebras, which are algebras where the product also has adjoint
operations which results in the Kleene star being more easily
axiomatized \cite{KOZEN1994366}.

@@

\begin{lemma}[\cite{day1970construction}]
  Under the assumptions above, the presheaf category $\Set^{\cat{C}}$ is
  monoidal closed.
\end{lemma}

@@

With this in mind, we want to design a dependent type theory that
builds on top of Kleene categories with the goal of internalizing
various reasoning principles for languages and their parsers. This
leads us to the abstract notion of model we are interested in
capturing with our type theory: a \emph{Chomsky category}.

\begin{definition}
  A Chomsky category is a locally Cartesian category with two hierarchies of
  universes $\{L_i\}_{i\in \nat}$ and $\{U_i\}_{i\in \nat}$ such that
  every $L_i$ and $U_i$ are $U_{i+1}$-small. Furthermore, we require
  $U_i$ to be closed under dependent products and sum,
  $L_i$ to be closed under the Kleene category connectives,
  dependent products, left and right closed structures, with
  a type constructor $G : L_i \to U_i$ and a linear dependent sum
  going the other direction.
\end{definition}

\begin{theorem}
  The presheaf category $\Set^{\cat{\Sigma^*}}$ is a Chomsky category.
\end{theorem}

Further, the syntactic category of the theory in \cref{sec:tt} is manifestly a Chomsky category.
\steven{This is true, right? That's like the entire point of having the syntax}

@@

Omission of the structural inference rules of a deductive system, such as in
linear logic \cite{GIRARD19871}, offers precise control over the number of times
a value may be used in a derivation. This control enables
\emph{resource-sensitive} reasoning, where we may treat a resource as
\emph{consumed} after usage. This viewpoint is amenable to parsing applications,
as we may treat characters of a string finite resources that are consumed at
parse-time. One may then envision a linear type system where the types comprise
formal grammars generated over some alphabet $\Sigma$, and the type constructors
correspond precisely to inductive constructions on grammars --- such as
conjunction, disjunction, and concatenation.

Unfortunately, programming in a linear term language is often cumbersome and
unintuitive for users. Code in such a language can become unnecessarily verbose
when using Girard encodings to translate high-level reasoning into low-level
linear terms. To alleviate this concern, in 1995 Benton et al.\ gave categorical
models of a logic that synthesized linear and non-linear logic
\cite{bentonMixedLinearNonlinear1995} --- appropriately referred to as a
\emph{linear-non-linear} (LNL) system. This work is simply typed, so the
boundary between linear and non-linear subtheories is entirely characterized via
a monoidal adjunction $F \dashv G$ between linear terms and non-linear terms.

\steven{Maybe there is more to say here, or perhaps later, that the only thing
  to gain from a non-dependent non-linear fragment are the things that are
  definiable via $G$. Except that these are not interesting, as they are
  necessarily empty}

\steven{Are these paragraphs recounting too much history?}

Inside of an LNL system, linearity may be thought of as an option that users can
choose to deliberately invoke at deliberate points in their developments in an
otherwise intuitionistic system. However, if we are wishing to treat parsers as
linear terms over input strings, the non-linear fragment of an LNL theory does
not really assist in the development of parsers. It is instead the case that
parsers may benefit from a \emph{dependence} on non-linear terms.

\steven{Mention sepration logic at some point. In particular the separating conjunction and magic wand}

Krishnaswami et al.\ equipped LNL systems with a restricted sense of dependent
types \cite{krishnaswami_integrating_2015}. When defining dependent types in
this setting dependence on linear terms is disallowed, but either linear or
non-linear terms are allowed to depend on non-linear terms.

The non-linear
subtheory corresponds to Martin-L\"of type theory (MLTT), which can be taken
as a foundation for mathematics. Dependence of linear terms on non-linear terms
of this form effectively allows us to bring any ordinary definitions from math
into scope as indicies and then define grammars relative to those indices. That
is all to say, the LNL dependent types in the style of Krishnaswami et al.\
allows us to recover the defintion of Aho's indexed grammars \cite{AhoIndexed} internal to our
type theory.
\steven{The above paragraph should maybe be put later. Also flows weirdly}

In this paper, we instantiate an LNL theory with dependent types as a deductive
setting to conduct parsing and formal grammar theory. In this system, linear
types encode formal grammars built over an alphabet/(set of base types)
$\Sigma$. The non-linear types are just MLTT/math defs. The syntax is in the
following subsections

\steven{Rephrase the above paragraph to summarize the intro of this section, as
  well as preview the syntax that's to come in the following subsections.}

\steven{Everything below here hasn't really been touched}

\subsection{Syntax}

@@

The structrual judgements are shown in \cref{fig:structjdg}, the typing
well-formedness in \cref{fig:typewf}, and the intuitionistic typing rules in
\cref{fig:inttyping}. There are all just as they appear in
\cite{krishnaswami_integrating_2015}.

@@

\begin{figure}
  \begin{mathpar}
    \boxed{\Gamma \vdash X : U_{i}}

    \inferrule{~}{\Gamma \vdash U_i : U_{i+1}}
 %
    \and
%
    \inferrule{~}{\Gamma \vdash L_i : U_{i+1}}
%
    \\
%
    \inferrule{\Gamma \vdash X : U_i \\ \hspace{-0.1cm} \Gamma, x : X \vdash Y : U_i}{\Gamma \vdash \PiTy x X Y : U_i }%
%
    \and
%
    \inferrule{\Gamma\vdash X : U_i \\ \hspace{-0.1cm} \Gamma, x : X \vdash Y : U_i}{\Gamma \vdash \SigTy x X Y : U_i}
%
    \\
%
    \inferrule{~}{\Gamma \vdash 1 : U_i}
%
    \and
%
    \inferrule{\Gamma \vdash A : L_i}{\Gamma \vdash G A : U_i}
%
    \\
%
    \inferrule{~}{\Gamma \vdash I : L_i}
 %
    \and
%
    \inferrule{\Gamma \vdash A : L_i \\ \hspace{-0.1cm}\Gamma \vdash B : L_i}{\Gamma \vdash A \otimes B : L_i}
%
    \and
%
    \inferrule{\Gamma \vdash A : L_i \\ \hspace{-0.1cm}\Gamma \vdash B : L_i}{\Gamma \vdash A \lto B : L_i}
%
    \and
%
    \inferrule{\Gamma \vdash A : L_i \\ \hspace{-0.1cm}\Gamma \vdash B : L_i}{\Gamma \vdash B \tol A : L_i}
%
    \\
%
    \inferrule{\Gamma \vdash X : U_i \\ \Gamma, x : X \vdash A : L_i}{\Gamma \vdash \LinPiTy x X A : L_i}
%
    \and
%
    \inferrule{\Gamma \vdash X : U_i \\ \Gamma, x : X \vdash A : L_i}{\Gamma \vdash \LinSigTy x X A : L_i}
%
    \\
%
    \inferrule{\Gamma \vdash X : U_i \quad \{\Gamma \vdash e_i : X\}_i}{\Gamma \vdash e_1 =_X e_2 : U_i}
    %
    \and
    %
    \inferrule{~}{\Gamma \vdash \top : L_i}
%
    \and
%
    \inferrule{\Gamma \vdash A : L_i \quad \Gamma \vdash B : L_i}{\Gamma \vdash A \amp B : L_i}
%
    \\
    %
    \inferrule{c \in \Sigma}{\Gamma \vdash c : L_0}
    %
    \and
    %
    \inferrule{\Gamma, x : L_i \vdash A : L_i \and A \textrm{ strictly positive}}{\Gamma \vdash \mu x.\, A : L_i}
  \end{mathpar}
  \caption{Type well-formedness}
  \label{fig:typewf}
\end{figure}

@@

    \inferrule{\Gamma \vdash p : e =_X e'}{\Gamma \vdash e \equiv e' : X}
%
    \and
%
    \inferrule{~}{\Gamma \vdash \app {(\lamb x e)} {e'} \equiv \subst x e {e'} : X}
%
    \and
%
    \inferrule{~}{\Gamma \vdash e \equiv \lamb x {\app e x} : \PiTy x X Y}
%
    \and
%
    \inferrule{~}{\Gamma \vdash \pi_1\, (e_1, e_2) \equiv e_1 : X}
%
    \and
%
    \inferrule{~}{\Gamma \vdash \pi_2\, (e_1, e_2) \equiv e_2 : \subst x {e_1} Y}
%
    \and
%
    \inferrule{~}{\Gamma \vdash e \equiv (\pi_1\, e, \pi_2\, e) : \SigTy x X Y}
%
    \and
%
    \inferrule{~}{}
%
    \inferrule{~}{\Gamma \vdash t \equiv t' : 1}
%
    \and
%
    \inferrule{~}{\Gamma \vdash G\, (G^{-1} \, t) \equiv t : G A}

@@

    \inferrule{~}{\Gamma; \cdot \vdash G^{-1}\, (G \, t ) \equiv t: A}
%
    \\
%
    \inferrule{~}{\Gamma; \Delta \vdash \app {(\lamblto a e)} {e'} \equiv \subst e x {e'} : C}
%
    \and
%
    \inferrule{~}{\Gamma; \Delta \vdash e \equiv \lamblto a {\app e a} : A \lto B}
%
    \and
%
    \inferrule{~}{\Gamma; \Delta \vdash \app {(\lambtol a e)} {e'} \equiv \subst e x {e'} : C}
%
    \and
%
    \inferrule{~}{\Gamma; \Delta \vdash e \equiv \lambtol a {\app e a} : A \tol B}
%
    \and
%
    \inferrule{~}{\Gamma; \Delta \vdash \app {(\dlamb x e)} {e'} \equiv \subst x a {e'} : C}
%
    \and
%
    \inferrule{~}{\Gamma; \Delta \vdash e \equiv \dlamb x {\app e x} : \LinPiTy x X A}
%
    \and
%
    \inferrule{~}{\Gamma; \Delta \vdash e \equiv e' : \top}
%
    \and
%
    \inferrule{~}{\Gamma; \Delta \vdash e_i \equiv \pi_i (e_1, e_2) : A_i}
%
    \and
%
    \inferrule{~}{\Gamma; \Delta \vdash e \equiv (\pi_1 e, \pi_2 e) : A\& B}
%
    \and
%
    \inferrule{~}{\Gamma; \Delta \vdash \letin {()} {()} e \equiv e : C}
    %
    \and
    %
    \inferrule{~}{\Gamma; \Delta \vdash \letin {()} e {\subst {e'} {()} a} \equiv \subst {e'} a e : C}
%
    \and
%
    \inferrule{~}{\Gamma; \Delta \vdash \letin {e \otimes e'} {a \otimes a'} e'' \equiv \subst {e''} {a, a'} {e, e'} : C}
%
    \and
    %
    \inferrule{~}{\Gamma; \Delta \vdash \letin {a \otimes b} e {\subst {e'} {a \otimes b} c} \equiv \subst {e'} c e : C}
%
    \and
%
    \inferrule{~}{\Gamma;\Delta \vdash \letin {(x, a)} {(e, e')} {e''} \equiv \subst {e''} {x, a} {e, e'} : C}
    %
    \and
    %
    \inferrule{~}{\Gamma; \Delta \vdash \letin {(x, a)} e {\subst {e'} {(x, a)} y} \equiv \subst {e'} e y : C}
\end{mathpar}
  \caption{Judgmental equality}
  \label{fig:jdgeq}
\end{figure}
\subsection{Syntax}
\label{subsec:syntax}

@@

\section{Semantics}
\label{sec:semantics}
\subsection{Semantics in Grammars}
\label{subsec:seming}
As mentioned earlier, the interpretation of grammars as functions from
$\String$ to $\Set$ serves as our intended semantics. When designing the syntax of this formalism, we were often
informed by this model. That is, there is an interpretation
$\llbracket \cdot \rrbracket$ of our syntax that provides grammar-based
semantics.

@@

\section{Automata as Grammars}
\label{sec:automata}
Classically, formal language theory is closely related to the study of automata.
In the Chomsky hierarchy, each language class is associated to a class of
automata that serve as recognizers. Internal to the type theory of \cref{sec:tt},
we can characterize these language classes syntactically; moreover, we
demonstrate the equivalence of these language classes to its associated automata class as a
proof term within our logic.

@@

\section{Related work}
\label{sec:related}
